Value-Suppressing Uncertainty Maps

### 2017-09-14 paper/main.tex
Revise / comment VSUP design section.

\begin{enumerate}
	\item There is a limited \emph{budget} of potential outputs in a bivariate map.
	\item Differences among \emph{certain} data are more significant than differences among \emph{uncertain} data.
\end{enumerate}

@@

However, if both of these assumptions hold, then it follows that the designer of a bivariate map should allocate more mark types to certain values, and fewer mark types to uncertain values. VSUPs codify this decision by \emph{reducing the number of mark categories for representing value as uncertainty increases.} This means that data encoded using a VSUP will make visible only the largest of differences in uncertain data, but highlight comparatively small differences in value when uncertainty is low. VSUPs therefore act as both a filtering mechanism (in that values with too much uncertainty are all mapped to the same glyph), as well as an implicit test of effect size (in that smaller and smaller changes in value are visible as uncertainty decreases). This strategy of dampening the low quality or high uncertainty values in maps in order to focus on more informative regions has measurable benefits, including the removal of statistically spurious visual patterns, and the highlighting of regions of interest~\cite{correll2017surprise}.

VSUPs rely on an underlying \textbf{quantization tree} that governs how values are discretized. Maximally uncertain values are all mapped to a singular ``root'' node. As uncertainty decreases, the tree branches into leaves, which evenly divide the data domain. The data value of a parent is the midpoint of all of its children. These leaves can then branch again, until a designer-specified stopping point. The layers act to the quantize the uncertainty domain, and the nodes the data domain. \figref{fig:example}, for example, is a VSUP with a tree with a branching factor of 2, and 4 layers, resulting in a total of 15 different output colors.

Correll \ea~\cite{correll2015layercake,correll2011visualizing} use a precursor of VSUPs in their LayerCake genomics visualization tool, where marks representing genomic data with increasing uncertainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into ``confidence fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection to uncertainty, such as saturation, value, blur, or size, also have deleterious effects on the disambiguation of colors and shapes. In both cases, the property of aliasing is \emph{ad hoc}, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUPs makes the choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.

Javascript code for generating VSUPs for use in D3 charts is available at \url{URL-REMOVED-FOR-REVIEW}.

@@

There are multiple choices that designers must make before creating a VSUP. In particular, they must choose 1) which visual channels map to value and uncertainty, 2) structure of the underlying VSUP tree.

Often, the graphical conventions of a chart reserve variables such as position and area for other uses, and determine the visual channel to be used for value. In heatmaps, for instance, value is regularly encoded using color, while vertical and horizontal position are reserved for location. In this paper, we primarily deal with the case where value is encoded using color. While there are some similar conventions for representing uncertainty, generally designers have more freedom in choosing a visual channel for uncertainty~\cite{maceachren1992visualizing}. We recommend channels with both a strong semiotic connection to uncertainty as well as a relatively large number of perceptually distinguishable levels. Here, we focus on value/saturation (or, equivalently, transparency against a white background). Correll \& Gleicher~\cite{correll2013error} show that even audiences without statistical backgrounds can interpret uncertainty information encoded in these channels.

@@

>>>>

\begin{enumerate}
	\item There is a limited \emph{budget} for perceptual discriminability.
	\item Differences among \emph{certain} data are more germane than differences among \emph{uncertain} data.
\end{enumerate}

@@

However, if both of these assumptions hold, then it follows that the designer of a bivariate map should allocate more mark types to certain values, and fewer mark types to uncertain values. VSUPs codify this decision by \emph{reducing the number of mark categories for representing value as uncertainty increases.} This means that data encoded using a VSUP will make visible only the largest of differences in uncertain data, but highlight comparatively small differences in value when uncertainty is low. VSUPs therefore act as both a filtering mechanism (in that values with too much uncertainty are mapped to the same glyph), as well as an implicit test of effect size (in that smaller and smaller changes in value are visible as uncertainty decreases). This strategy of dampening low quality or highly uncertain values in maps in order to focus on more informative regions has measurable benefits, including the removal of statistically spurious visual patterns, and the highlighting of regions of interest~\cite{correll2017surprise}.

VSUPs rely on an underlying \textbf{quantization tree} that governs how values are discretized. Maximally uncertain values are mapped to a singular ``root'' node. As uncertainty decreases, the tree branches into leaves, which evenly divide the data domain. The data value of a parent is the midpoint of all of its children. These leaves can then branch again, up to a designer-specified stopping point. The layers quantize the uncertainty domain, and sibling nodes quantize the data domain. \figref{fig:example} shows a VSUP with a tree with a branching factor of 2 and 4 layers, resulting in an output range with 15 distinct colors.

Correll \ea~\cite{correll2015layercake,correll2011visualizing} use a precursor of VSUPs in their LayerCake genomics visualization tool, where marks representing genomic data with increasing uncertainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into a ``confidence fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection to uncertainty, such as saturation, value, blur, or size, also have deleterious effects on the disambiguation of colors and shapes. In both cases, the property of aliasing is \emph{ad hoc}, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUPs makes the choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.

\textbf{JH: The precending paragraph might be better placed in related work!}

JavaScript code for generating VSUPs for use in D3 charts is available at \url{URL-REMOVED-FOR-REVIEW}.

@@

There are multiple choices that designers must make before creating a VSUP. Specifically, they must choose (1) which visual channels map to value and uncertainty and (2) the structure of the underlying VSUP tree.

Often, the graphical conventions of a chart reserve variables such as position and area for other uses, and determine the visual channel to be used for value. In heatmaps, for instance, value is regularly encoded using color, while vertical and horizontal position are reserved for location. \textbf{JH: This is the third time we've pointed this out...} In this paper, we primarily deal with the case where value is encoded using color. While there are some similar conventions for representing uncertainty, generally designers have more freedom in choosing a visual channel for uncertainty~\cite{maceachren1992visualizing}. \textbf{This feels like we are revisiting introductory and related work material again!} We recommend channels with both a strong semiotic connection to uncertainty as well as a relatively large number of perceptually distinguishable levels. Here, we focus on value/saturation (or, equivalently, transparency against a white background). Correll \& Gleicher~\cite{correll2013error} provide evidence that even audiences without statistical training can successfully interpret uncertainty information encoded in these channels.

\textbf{We should say something about the choice of value color map. Generally speaking, single-hue luminance ramps must be used with caution if they ramp too strongly towards white. This will be confused with manipulations of opacity or saturation. Viridis might be attractive due to its avoidance of desaturated colors.}

@@

\textbf{A few thoughts about this section based on the current figures. (1) Fig. 3 in particular has two graphics that look pretty similar. Is there anything we should do to draw attention to the most salient differences between conditions? At minimum, we have the text, but we could consider annotations or even a third vis that is a heatmap of differences between charts. (2) A third example using a different visual form might be useful. For example, a node-link network with value + uncertainty associated with each node.}
